---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 300,
  fig.width = 7
)
```

# structr

<!-- badges: start -->

[![R-CMD-check](https://github.com/tobiste/structr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tobiste/structr/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

`{structr}` is a free and open-source package for R that provides tools for structural geology. The toolset includes

-   Analysis and visualization of orientation data of structural geology (including, **stereographic projections**, contouring, fabric plots, and statistics),

-   Statistical analysis: spherical mean and variance, confidence regions, hypothesis tests, **cluster analysis** of orientation data (`sph_cluster()`, and geodesic regression to find the **best-fitting great circle or small circle** through orientation data (`regression_greatcircle()` and `regression_smallcircle()`),

-   Reconstruction of fabric orientations in **oriented drillcores** by transforming the α, β, and γ angles (`drillcore_transformation()`,

-   **Stress analysis**: reconstruction of stress orientation and magnitudes from fault-slip data (stress inversion based on **Michael, 1984**: `slip_inversion()`), extracting the **maximum horizontal stress** of a 3D stress tensor (`SH()`), and visualization of magnitudes of stress in the **Mohr circle** (`Mohr_plot()`),

-   Calculation fault displacement components,

-   Strain analysis (**R**<sub>f</sub>/ϕ), contouring on the unit hyperboloid, **Fry plots** and **Hsu plots**

-   Vorticity analysis using the **Rigid Grain Net** method (`RGN_plot()`), and

-   Direct import of your field data from **StraboSpot** projects (`read_strabo_JSON()`).

> The {structr} package is all about structures in 3D. For analyzing orientations in 2D (statistics, rose diagrams, etc.), check out the [tectonicr](https://github.com/tobiste/tectonicr) package!

## Installation

You can install the development version of `{structr}` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("tobiste/structr")
```

## Documentation

The detailed documentation can be found at <https://tobiste.github.io/structr/>

## Examples

These are some basic examples which shows you what you can do with {structr}. First we load the package

```{r load, warning=FALSE,message=FALSE}
library(structr)
```

### Stereographic and equal-area projection

Plot orientation data in equal-area, lower hemisphere projection:

```{r stereo, warning=FALSE,message=FALSE}
data("example_planes")
data("example_lines")

par(xpd = NA)
stereoplot(title = "Lambert equal-area projection", sub = "Lower hemisphere")
points(example_lines, col = "#B63679", pch = 19, cex = .5)
points(example_planes, col = "#000004", pch = 1, cex = .5)

legend("topright", legend = c("Lines", "Planes"), col = c("#B63679", "#000004"), pch = c(19, 1), cex = 1)
```

### Density

Density shown by contour lines...

```{r stereo_density_lines, warning=FALSE,message=FALSE}
par(mfrow = c(1, 2))
contour(example_planes)
points(example_planes, col = "grey", cex = .5)
title(main = "Planes")

contour(example_lines)
points(example_lines, col = "grey", cex = .5)
title(main = "Lines")
```

or as filled contours:

```{r stereo_density_img, warning=FALSE,message=FALSE}
par(mfrow = c(1, 2))
image(example_planes)
points(example_planes, col = "grey", cex = .5)
title(main = "Planes")

image(example_lines)
points(example_lines, col = "grey", cex = .5)
title(main = "Lines")
```

### Spherical statistics

Calculation of arithmetic mean, geodesic mean, confidence cones and eigenvectors... and plotting them in the equal-area projection:

```{r stats, warning=FALSE,message=FALSE}
planes_mean <- sph_mean(example_planes)
planes_geomean <- geodesic_mean(example_planes)
planes_eig <- ot_eigen(example_planes)$vectors

par(mfrow = c(1, 2), xpd = NA)
stereoplot(title = "Planes", guides = FALSE)
points(example_planes, col = "lightgrey", pch = 1, cex = .5)
lines(planes_eig, col = c("#FB8861FF", "#FEC287FF", "#FCFDBFFF"), lty = 1:3)
points(planes_mean, col = "#B63679", pch = 19, cex = 1)
points(planes_geomean, col = "#E65164FF", pch = 19, cex = 1)
points(planes_eig, col = c("#FB8861FF", "#FEC287FF", "#FCFDBFFF"), pch = 19, cex = 1)
legend(
  0, -1.1,
  xjust = .5,
  legend = c("Arithmetic mean", "Geodesic mean", "Eigen 1", "Eigen 2", "Eigen 3"),
  col = c("#B63679", "#E65164FF", "#FB8861FF", "#FEC287FF", "#FCFDBFFF"),
  pch = 19, lty = c(NA, NA, 1, 2, 3),
  cex = .75
)

lines_mean <- sph_mean(example_lines)
lines_delta <- delta(example_lines)
lines_confangle <- confidence_ellipse(example_lines)

stereoplot(title = "Lines", guides = FALSE)
points(example_lines, col = "lightgrey", pch = 1, cex = .5)
points(lines_mean, col = "#B63679", pch = 19, cex = 1)
stereo_confidence(lines_confangle, col = "#E65164FF")
lines(lines_mean, ang = lines_delta, col = "#FB8861FF")
legend(
  0, -1.1,
  xjust = .5,
  legend = c("Arithmetic mean", "95% confidence cone", "63% data cone"),
  col = c("#B63679", "#E65164FF", "#FB8861FF"),
  pch = c(19, NA, NA), lty = c(NA, 1, 1), cex = .75
)
```

### Orientation tensor and fabric plots

The shape parameters of the orientation tensor of the above examples planes and lines can be visualized in two ways:

```{r stereo_ortensor, warning=FALSE,message=FALSE}
par(mfrow = c(1, 2), xpd = NA)
vollmer_plot(example_planes, col = "#000004", pch = 16)
vollmer_plot(example_lines, col = "#B63679FF", pch = 16, add = TRUE)
title("Fabric plot of Vollmer (1990)")

hsu_plot(example_planes, col = "#000004", pch = 16)
hsu_plot(example_lines, col = "#B63679FF", pch = 16, add = TRUE)

legend(
  2.5, -.25,
  xjust = .5, horiz = TRUE, xpd = NA,
  legend = c("Planes", "Lines"), col = c("#000004", "#B63679FF"), pch = 16
)
```

#### Best-fit great and small-circles (geodesic regression)

Finds the best-fit great or small-circle for a given set of vectors by applying geodesic regression:

```{r bestfit, warning=FALSE,message=FALSE}
set.seed(20250411)
data("gray_example")
cleavage <- gray_example[1:8, ]
bedding <- gray_example[9:16, ]

cleavage_gc <- regression_greatcircle(cleavage)
bedding_gc <- regression_greatcircle(bedding)

cleavage_sc <- regression_smallcircle(cleavage)
bedding_sc <- regression_smallcircle(bedding)

par(mfrow = c(1, 2), xpd = NA)
stereoplot(title = "Best greatcircle", guides = FALSE)
lines(cleavage_gc$vec, col = "#000004FF")
lines(bedding_gc$vec, col = "#B63679")
points(cleavage, col = "#1D1147")
points(bedding, col = "#E65164", pch = 4)

legend(
  0, -1.1,
  xjust = .5,
  col = c("#000004FF", "#B63679"),
  lty = c(1, 1), legend = c("Cleavage greatcircle", "Bedding greatcircle"), bg = "white"
)

stereoplot(title = "Best smallcircle", guides = FALSE)
lines(cleavage_sc$vec, cleavage_sc$cone, col = "#000004FF")
lines(bedding_sc$vec, bedding_sc$cone, col = "#B63679")
points(cleavage, col = "#1D1147")
points(bedding, col = "#E65164", pch = 4)

legend(0, -1.1,
  xjust = .5,
  col = c("#000004FF", "#B63679"), lty = c(1, 1), legend = c("Cleavage smallcircle", "Bedding smallcircle"), bg = "white"
)
```

### Fault plots

Graphical representation of fault-slip data using Angelier plot (slip vector on fault plane great circle) and Hoeppener plot (fault slip vector projected on pole to fault plane):

```{r stereo_faults, warning=FALSE,message=FALSE}
data("angelier1990")
faults <- angelier1990$TYM

par(mfrow = c(1, 2))
stereoplot(title = "Angelier plot")
angelier(faults)

stereoplot(title = "Hoeppener plot")
hoeppener(faults, points = FALSE)
```

### Fault-slip inversion

Compute deviatoric stress tensor and calculate 95% confidence intervals using bootstrap samples:

```{r stereo_inversion1, warning=FALSE,message=FALSE}
set.seed(20250411)
faults_stress <- slip_inversion(faults, boot = 10)
```

Visualize the slip inversion results (orientation of principal stresses):

```{r stereo_inversion_plot, warning=FALSE,message=FALSE}
cols <- c("#000004FF", "#B63679FF", "#FEC287FF")
R_val <- round(faults_stress$R, 2)
R_CI <- round(faults_stress$R_conf, 2)

stereoplot(
  title = "Principal stress axes",
  sub = paste0("Relative stress magnitudes R = ", R_val, " | ", "95% CI: [", R_CI[1], ", ", R_CI[2], "]"),
  guides = FALSE
)
angelier(faults, col = "grey80")
stereo_confidence(faults_stress$principal_axes_conf$sigma1, col = cols[1])
stereo_confidence(faults_stress$principal_axes_conf$sigma2, col = cols[2])
stereo_confidence(faults_stress$principal_axes_conf$sigma3, col = cols[3])
text(faults_stress$principal_axes,
  label = rownames(faults_stress$principal_axes),
  col = cols, adj = -.25
)
```

Visualize the accuracy of the slip inversion by showing the deviation angle (&beta;) between the theoretical slip and the actual slip vector:

```{r stereo_inversion_deviation, warning=FALSE,message=FALSE}
beta <- faults_stress$fault_data$beta
beta_mean <- round(faults_stress$beta)
beta_CI <- round(faults_stress$beta_CI)

stereoplot(
  title = "Stress inversion accuracy",
  sub = bquote("Average deviation" ~ bar(beta) == .(beta_mean) * degree ~ "\U00B1" ~ .(beta_CI) * degree),
  guides = FALSE
)
angelier(faults, col = assign_col(beta))
legend_c(
  seq(min(beta), max(beta), 10),
  title = bquote("Deviation angle" ~ beta ~ "(" * degree * ")")
)
```

Azimuth of the maximum horizontal stress (in degrees) for the slip inversion result:

```{r stereo_inversion_SH, warning=FALSE,message=FALSE}
# Simply call
# faults_stress$SHmax
# faults_stress$SHmax_CI # confidence interval

SH(
  S1 = faults_stress$principal_axes[1, ],
  S2 = faults_stress$principal_axes[2, ],
  S3 = faults_stress$principal_axes[3, ],
  R = faults_stress$R
)
```

### Mohr circles

The Mohr circle for the slip inversion result:

```{r stereo_inversion_mohr, warning=FALSE,message=FALSE}
Mohr_plot(
  sigma1 = faults_stress$principal_vals[1],
  sigma2 = faults_stress$principal_vals[2],
  sigma3 = faults_stress$principal_vals[3],
  unit = NULL, include.zero = FALSE
)
points(faults_stress$fault_data$sigma_n, abs(faults_stress$fault_data$sigma_s),
  col = assign_col(beta), pch = 16
)
```

### Strain analysis

#### 2D Strain

Aspect ratio of finite strain ellipses vs orientation of long-axis (Rf/ϕ)

```{r strain, warning=FALSE,message=FALSE}
data(ramsay)

par(mfrow = c(1, 2))
Rphi_plot(r = ramsay[, 1], phi = ramsay[, 2])
elliott_plot(ramsay[, 1], ramsay[, 2], proj = "eqd")
```

#### 3D Strain

Finite strain ellipsoids plotted in Flinn diagram and Hsu diagram:

```{r strain3D, warning=FALSE,message=FALSE}
data("holst")
R_XY <- holst[, "R_XY"]
R_YZ <- holst[, "R_YZ"]

par(mfrow = c(1, 2))
flinn_plot(cbind(R_XY, R_YZ), log = TRUE, col = "#B63679", pch = 16)
hsu_plot(cbind(R_XY, R_YZ), col = "#B63679", pch = 16)
```

### Vorticity analysis

Aspect ratio of finite strain ellipses of porphyroclasts vs orientation of long-axis with respect to foliation plotted in the **Rigid Grain Net**

```{r rgn, warning=FALSE,message=FALSE}
data(shebandowan)
set.seed(20250411)

# Color code porphyroclasts by size of clast (area in log-scale):
RGN_plot(shebandowan$r, shebandowan$phi, col = assign_col(log(shebandowan$area)), pch = 16)
```

## Author

Tobias Stephan ([tstephan\@lakeheadu.ca](mailto:tstephan@lakeheadu.ca){.email})

## Feedback, issues, and contributions

I welcome feedback, suggestions, issues, and contributions! If you have found a bug, please file it [here](https://github.com/tobiste/structr/issues) with minimal code to reproduce the issue.

## License

MIT License
