---
title: "1. Basics"
author: "Tobias Stephan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Basics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

 
```{r setup, warning=FALSE, message=FALSE}
library(structr)
```

## Import 

```{r read_csv, echo=FALSE, eval=FALSE}
read.table("path/to/my/file.txt", header = TRUE, sep = "\t")
read.csv("path/to/my/file.csv")
readxl::read_xlsx("path/to/my/file.xlsx")
```

From a Strabospot project, download the json file and import it using the `read_strabo_JSON()` function.

 
```{r read_json, echo=FALSE, eval=FALSE}
read_strabo_JSON("path/to/my/file.json")
```

This returns a list which contains all the information and metadata (including coordinates, descriptions etc) extracted from the Strabospot project in the list element `data`, the tags used for the project in the list element `tags`, and the linear and planar orientation measurements in the list elements `lines` and `planes`, respectively.

 
## Data conversion to spherical objects
Usually orientation data is stored in a table containing the column dip direction (or strike) and the dip angle of a measured plane...

 
```{r data_p}
data(example_planes)
head(example_planes)
```

 
or the trend (azimuth) and plunge (inclination) of a measured line...

```{r data_l}
data(example_lines)
head(example_lines)
```

To convert these data frames to spherical objects, use the `Plane()` and `Line()` functions from the `structr` package. These functions take the dip direction and dip angle for planes, and the trend and plunge for lines as arguments.

```{r data_conversion}
data(example_planes)
planes <- Plane(example_planes$dipdir, example_planes$dip)
lines <- Line(example_lines$trend, example_lines$plunge)
```
 > If the raw data was imported using `read_strabo_JSON()` this step is not necessary as the data will come already in the correct format.

The spherical objects can be easily converted into Cartesian coordinate vectors using the functions `line2vec()` or `plane2vec()`:

```{r data_transformation}
lines_vector <- Vec3(lines)
head(lines_vector)
```

 
```{r data_transformation2}
planes_as_line <- Vec3(planes)
head(planes_as_line)
```

which produces the same output as

 
```{r data_transformation3}
head(Vec3(planes_as_line))
```

 
## Some helpers
Converts strike into dip direction using right-hand rule

```{r strike2dipdir}
rhr2dd(271)
dd2rhr(1)
```

```{r helpers}
# parse_strike_dip(c("E30N", "W10S"), c("45NW", "4SE"))
```

 
## Vector operations
Since the spherical or vector objects are easily convertible, they can be used for all sort of vector operations, such as the magnitude (or length), 
the angle between vectors, dot product, cross product, projection and rotation.

Define some example vectors:

```{r math0}
line1 <- Line(120, 50)
line2 <- Line(10, 30)
```

 
The **vector length** (or magnitude):

```{r math_norm}
vlength(line1)
```

> Orientation vectors are by definition unit vectors, i.e. their length is equal to 1.

The **angle** between two vectors

```{r math_angle}
angle(line1, line2)
```

 
The **dot product** (or scalar product) of two  vectors

```{r math_dot}
line1 %*% line2
```

Intuitively, it tells us something about how much two vectors point in the same direction.

The **cross product** of two  vectors:

 
```{r math_cross}
crossprod(line1, line2)
```

This gives the vector that is perpendicular to the plane spanned by the two vectors.

The **projection** of a vector on another vector:

```{r math_project}
project(line1, line2)
```

> Because of vectors a unit vectors, the projected vector is equal to the second vector.

 
The **rotation** of a vector about another vector (rotation axis) by a specified rotation angle:

```{r math_rotate}
rotate(line1, rotaxis = line2, rotangle = 45)
```
