---
title: "Vector Algebra"
author: "Tobias Stephan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
description: >
  This vignette explains the basic objects used in the package.
vignette: >
  %\VignetteIndexEntry{Vector algebra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

This tutorial introduces the basic mathematical operations of the `{structr}` package for analyzing orientation data.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

 
```{r setup, warning=FALSE, message=FALSE}
library(structr)
```


## Vector operations
Since the spherical or vector objects are easily convertible, they can be used for all sort of vector operations, such as the magnitude (or length), 
the angle between vectors, dot product, cross product, projection and rotation.

Define some example vectors:

```{r math0}
line1 <- Line(120, 50)
line2 <- Line(10, 30)
```

 
The **vector length** (or magnitude):

```{r math_norm}
vector_length(line1)
```

> Orientation vectors are by definition unit vectors, i.e. their length is equal to 1.

The **angle** between two vectors

```{r math_angle}
angle(line1, line2)
```

 
The **dot product** (or scalar product) of two  vectors

```{r math_dot}
dotprod(line1, line2)
```

Intuitively, the dot product tells us how much two vectors point in the same direction.

The **cross product** of two  vectors:

 
```{r math_cross}
crossprod(line1, line2)
```

This gives the vector that is perpendicular to the plane spanned by the two vectors.

The **projection** of a vector on another vector:

```{r math_project}
project(line1, line2)
```

> Because the vectors are both unit vectors, the projected vector is equal to the second vector.

 
The **rotation** of a vector about another vector (rotation axis) by a specified rotation angle:

```{r math_rotate}
rotate(line1, rotaxis = line2, rotangle = 45)
```

**Linear transformation** transforms vectors using a transformation matrix (second-order tensor). 

```{r math_transform}
trans_mat <- matrix(runif(9), 3, 3)
transform_linear(line1, trans_mat)
```

