---
title: "Faults and Paleostress"
author: "Tobias Stephan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
description: >
  Introduction into the Fault object, and how to perform stress inversions
vignette: >
  %\VignetteIndexEntry{Faults and Paleostress}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

This tutorial demonstrates the `"Fault"` object. It also shows how you
can extract paleo-stress directions from fault slip data.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(20250411)
```

## Fault objects

```{r setup, warning=FALSE, message=FALSE}
# load the structr package
library("structr")

# set up some plotting defaults
options(
  structr.upper.hem = FALSE, structr.earea = TRUE, 
  structr.guides = FALSE
  )
```

A fault is given by the orientation of its plane (dip direction and dip
angle), the orientation of the slip (e.g. measured from striae, given in
azimuth and plunge angles), and the sense of displacement:

```{r faults_import}
my_fault <- Fault(120, 50, 60, 110, sense = -1)
```

> Sense of fault displacement is 1 or -1 for normal or thrust offset,
> respectively.

Rake of the fault, i.e. the angle between fault slip vector and fault
strike:

```{r faults_rake1}
Fault_rake(my_fault)
```

Define a fault by just knowing the orientation of the fault plane, the
sense, and the rake

```{r faults_rake2}
# 1. Define a plane through dip direction, dip angle
fault_plane <- Plane(c(120, 120, 100), c(60, 60, 50))

# 2. Define a fault through the plane and rake angle:
Fault_from_rake(fault_plane, rake = c(84.7202, -10, 30))
```

Often, measured orientation angles can be (slightly) imprecise and
subjected to some random noise. Thus the slip vector will not lie
(perfectly) on the fault plane, judging by the measurements. To correct
the measurements so that this will not be the case:

```{r faults_correct}
p <- Pair(120, 60, 110, 58)
```

```{r faults_correct2}
misfit_pair(p)
```

```{r faults_correct3}
correct_pair(p)
```

> A `"Pair"` object is a container of associated plane and line
> measurements. Basically like a fault without the sense of
> displacement.

## Plotting faults

Fault objects consist of planes (fault plane), lines (e.g. striae), and
the sense of movement. There are two ways how these combined features
can be visualized, namely the Angelier and the Hoeppener plot.

-   The **Angelier plot** shows all planes as *great circles* and
    lineations as points (after Angelier, 1984). Fault striae are
    plotted as vectors on top of the lineation pointing in the movement
    direction of the hanging wall. Easy to read in case of homogeneous
    or small datasets.
-   The **Hoeppener plot** shows all planes as *poles* while lineations
    are not shown (after Hoeppener, 1955). Instead, fault striae are
    plotted as vectors on top of poles pointing in the movement
    direction of the hanging wall. Useful in case of large or
    heterogeneous datasets.

```{r faults_plotting}
#| fig.alt: >
#|   Diagram showing some fault data in equa-area projection using the Angelier
#|   and Hoeppener representation of fault-slip
# simongomez is a example fault dataset:

# define some colors for each fault in the dataset (here the fault sense)
fault_cols <- assign_col(simongomez[, 5], pal = viridis::magma, begin = .2, end = .8)

par(mfrow = c(1, 2))
stereoplot(title = "Angelier plot")
angelier(simongomez, col = fault_cols)

stereoplot(title = "Hoeppener plot")
hoeppener(simongomez, col = fault_cols, points = FALSE)
```

## Fault stress analysis

The *Wallace-Bott hypothesis* states that fault slip occurs parallel to
the maximum shear stress. This allows to reconstruct stress axes using
fault-slip data. {structr} offers several techniques to calculate the
orientation of principal stress axes, the simple P-T method, and a
fault-slip inversion technique.

### P-T method

This simple technique calculates PT-axes, kinematic plane (M), and
dihedra separation plane (d).

First we load some example data (here the first three faults from the
TYM dataset by from Angelier, 1990)[^1]

[^1]: Angelier, J. (1990). Inversion of field data in fault tectonics to
    obtain the regional stress—III. A new rapid direct inversion method
    by analytical means. Geophys. J. Int, 103, 363–376.
    <https://doi.org/10.1111/j.1365-246X.1990.tb01777.x>

```{r faults_pt1}
data("angelier1990")
my_fault2 <- angelier1990$TYM[1:3, ]

print(my_fault2)
```

```{r faults_pt2}
my_fault2_PT <- Fault_PT(my_fault2)
print(my_fault2_PT)
```

Plot the results

```{r faults_pt_plot}
#| fig.alt: >
#|   Diagram showing some fault-slip data and the PT axes in a stereoplot
stereoplot(title = "PT results")
fault_plot(my_fault2)
points(my_fault2_PT$p, col = "#B63679FF", pch = 16)
points(my_fault2_PT$t, col = "#FEC287FF", pch = 18)
lines(my_fault2_PT$t, lty = 2, col = "grey40")
lines(my_fault2_PT$d, lty = 3, col = "grey80")

legend("right",
  legend = c("P-axis", "T-axis", "M-plane", "Diheadra"),
  col = c("#B63679FF", "#FEC287FF", "grey40", "grey80"),
  pch = c(16, 18, NA, NA), lty = c(NA, NA, 2, 3)
)
```

### Fault slip inversion

Our goal is to find the single uniform stress tensor that most likely
caused the faulting events. With only slip data to constrain the stress
tensor the isotropic component can not be determined, unless assumptions
about the fracture criterion are made. Hence inversion will be for the
deviatoric stress tensor only. A single fault can not completely
constrain the deviatoric stress tensor a, therefore it is necessary to
simultaneously solve for a number of faults, so that a single a that
best satisfies all of the faults is found.

> This is equivalent to assuming that the stress field is a constant
> tensor within the region being studied for the duration of the
> faulting event.

{structr} provides four numerical solutions to determine the
orientation of the principal stresses from fault slip data.

-   *Michael (1984)*: Direct inversion method[^2] which uses
    bootstrapping for confidence intervals of the stress estimates.

-   *Angelier (1990)*: Direct inversion method[^3] coupled with the
    iterative optimization after Mostafa (2005)[^4] to find the best fit
    reduced stress tensor.

-   *Yamaji & Sato (2006)*: Direct inversion method using the
    5-dimensional parameter space[^5].

-   *Hansen (2013)*: Direct inversion using a 9-dimensional parameter
    space, useful when vorticity affects the fault-slip data[^6].

[^2]: Michael, A. J. (1984). Determination of stress from slip data:
    Faults and folds. Journal of Geophysical Research: Solid Earth,
    89(B13), 11517–11526. <https://doi.org/10.1029/JB089iB13p11517>

[^3]: Angelier, J. (1990). Inversion of field data in fault tectonics to
    obtain the regional stress—III. A new rapid direct inversion method
    by analytical means. Geophys. J. Int, 103, 363–376.
    <https://doi.org/10.1111/j.1365-246X.1990.tb01777.x>

[^4]: Mostafa, M. E. (2005). Iterative direct inversion: An exact
    complementary solution for inverting fault-slip data to obtain
    palaeostresses. Computers & Geosciences, 31(8), 1059–1070.
    <https://doi.org/10.1016/j.cageo.2005.02.012>

[^5]: Yamaji, A., & Sato, K. (2006). Distances for the solutions of
    stress tensor inversion in relation to misfit angles that accompany
    the solutions. Geophysical Journal International, 167(2), 933–942.
    <https://doi.org/10.1111/j.1365-246X.2006.03188.x>

[^6]: Hansen, J. A. (2013). Direct inversion of stress, strain or strain
    rate including vorticity: A linear method of homogenous fault-slip
    data inversion independent of adopted hypothesis. Journal of
    Structural Geology, 51, 3–13.
    <https://doi.org/10.1016/j.jsg.2013.03.014>

First we load some example data (here the data from Angelier, 1990)[^7]

[^7]: Angelier, J. (1990). Inversion of field data in fault tectonics to
    obtain the regional stress—III. A new rapid direct inversion method
    by analytical means. Geophys. J. Int, 103, 363–376.
    <https://doi.org/10.1111/j.1365-246X.1990.tb01777.x>

```{r slip_inversion1}
#| fig.alt: >
#|   Diagram showing some fault-slip data in a stereoplot
fault_data <- angelier1990$TYM

stereoplot(title = "Test data")
fault_plot(fault_data, col = "grey30")
```

The stress inversion using the Michael method with 10 bootstraps:

```{r slip_inversion_check}
inv_res <- slip_inversion(fault_data, method = "michael", n_iter = 10)

# Average alpha angle
inv_res$misfit$alpha

# Average resolved shear stress
inv_res$tau_mean
```

To check the accuracy of the solution, you can evaluate

-   the average angle β between the tangential traction predicted by the
    best stress tensor and the slip vector on each plane (ideally close
    to 0), and
-   the average resolved shear stress on each plane (should be close to
    1).
-   the average "Ratio Upsilon" (RUP) parameter after Angelier (1990),
    ranging from 0 (perfect fit) to 200% (misfit)

To visualizing the orientation of the principal stresses and the
confidence region of the axes, you may use the [stereoplot()] functions
from {structr}

```{r slip_inversion_plot}
#| fig.alt: >
#|   Diagram showing principal stress vector results in a stereoplot

cols <- c("#000004FF", "#B63679FF", "#FEC287FF")

stereoplot(title = "Stress inversion")
fault_plot(fault_data, col = "grey75")
stereo_confidence(inv_res$principal_axes_CI$sigma1, col = cols[1])
stereo_confidence(inv_res$principal_axes_CI$sigma2, col = cols[2])
stereo_confidence(inv_res$principal_axes_CI$sigma3, col = cols[3])
text(inv_res$principal_axes,
  label = rownames(inv_res$principal_axes),
  col = cols, adj = -.25
)
legend("topleft",
  col = cols,
  legend = rownames(inv_res$principal_axes), pch = 16
)
```

The stress shape ratio after Angelier (1979)[^8]:
$\Phi  = (\sigma_2 - \sigma_3)/(\sigma_1 - \sigma_3)$

[^8]: Angelier, J. (1979). Determination of the mean principal
    directions of stresses for a given fault population. Tectonophysics,
    56(3–4), T17–T26.
    [https://doi.org/10.1016/0040-1951(79)90081-7](https://doi.org/10.1016/0040-1951(79)90081-7){.uri}

```{r slip_inversion_phi}
inv_res$stress_shape$phi

# 95% confidence interval
inv_res$phi_CI
```

The angle α is the angle between the tangential traction predicted by
the best stress tensor and the slip vector. This deviation can be
visualized in the stereoplot:

```{r slip_inversion_alpha}
#| fig.alt: >
#|   Diagram showing deviation of the best-fit principal stress tensor in a
#|   stereoplot
alpha <- inv_res$misfit$alpha

stereoplot(
  title = "Deviation",
  sub = bquote(bar(alpha) == .(round(inv_res$alpha)) * degree)
)

fault_plot(fault_data, col = assign_col(alpha))
legend_col(
  seq(min(alpha), max(alpha), 10),
  title = bquote("Deviation angle" ~ alpha ~ "(" * degree * ")")
)
```

```{r slip_inversion_mohr}
#| fig.alt: >
#|   Diagram showing the Mohr circle, and the normal and shear stresses for the
#|   slip inversion result
stress_components <- tau2shearnorm(inv_res$stress_tensor, fault_data, friction = 0.6) 

Mohr_plot(
  sigma1 = inv_res$principal_vals[1],
  sigma2 = inv_res$principal_vals[2],
  sigma3 = inv_res$principal_vals[3],
  unit = NULL, include.zero = FALSE
)
points(stress_components[, 'normal'], abs(stress_components[, 'shear']),
  col = assign_col(alpha), pch = 16
)
```

### Maximum horizontal stress

The orientation of the maximum horizontal stress ($\sigma_\text{Hmax}$)
can be calculated from the stress tensor the the orientation of the
principal stress ($\sigma_1$, $\sigma_2$, $\sigma_3$) axes their their
relative magnitudes ($R = (\sigma_1 - \sigma_2)/(\sigma_1 - \sigma_3)$)
[^9].

[^9]: Lund & Townend (2007): Calculating horizontal stress orientations
    with full or partial knowledge of the tectonic stress tensor.
    *Geophys. J. Int.*, 170, 1328—1335. doi:
    10.1111/j.1365-246X.2007.03468.x

First, we define the orientation of the principle stress axes:

```{r lund0}
S1 <- Line(250.89, 70.07)
S3 <- Line(103.01, 17.07)
```

To get $\sigma_2$, which is perpendicular to $\sigma_1$ and $\sigma_3$,
we calculate the cross-product of the two vectors:

```{r lund01}
S2 <- crossprod(S3, S1)
```

The azimuth of $\sigma_\text{Hmax}$ for a given stress ratio `R = 1`:

```{r lund1}
SH(S1, S2, S3, R = 1) # in degrees
```

For a several stress ratios:

```{r lund2}
R <- seq(0, 1, .1)
cbind(R, SH = SH(S1, S2, S3, R = R))
```

The $\sigma_\text{Hmax}$ direction for our slip inversion result from
above:

```{r lund_3}
inv_shmax <- SH(
  S1 = inv_res$principal_axes[1, ],
  S2 = inv_res$principal_axes[2, ],
  S3 = inv_res$principal_axes[3, ],
  R = inv_res$stress_shape$R
)
print(inv_shmax)
```

This (or any) direction can be added as compression arrows to a stereoplot using [stereo_shmax()]:

```{r stereo_shmax}
stereoplot(title = "Stress inversion")
fault_plot(fault_data, col = "grey75")
stereo_confidence(inv_res$principal_axes_CI$sigma1, col = cols[1])
stereo_confidence(inv_res$principal_axes_CI$sigma2, col = cols[2])
stereo_confidence(inv_res$principal_axes_CI$sigma3, col = cols[3])
text(inv_res$principal_axes,
  label = rownames(inv_res$principal_axes),
  col = cols, adj = -.25
)

stereo_shmax(inv_shmax)
```

## References

Angelier, J. (1979). Determination of the mean principal directions of
stresses for a given fault population. Tectonophysics, 56(3–4), T17–T26.
[https://doi.org/10.1016/0040-1951(79)90081-7](https://doi.org/10.1016/0040-1951(79)90081-7){.uri}

Angelier, J. (1990). Inversion of field data in fault tectonics to
obtain the regional stress—III. A new rapid direct inversion method by
analytical means. Geophys. J. Int, 103, 363–376.
<https://doi.org/10.1111/j.1365-246X.1990.tb01777.x>

Hansen, J. A. (2013). Direct inversion of stress, strain or strain rate
including vorticity: A linear method of homogenous fault-slip data
inversion independent of adopted hypothesis. Journal of Structural
Geology, 51, 3–13. <https://doi.org/10.1016/j.jsg.2013.03.014>

Lund, B., & Townend, J. (2007). Calculating horizontal stress
orientations with full or partial knowledge of the tectonic stress
tensor. Geophysical Journal International, 170(3), 1328–1335.
<https://doi.org/10.1111/j.1365-246X.2007.03468.x>

Michael, A. J. (1984). Determination of stress from slip data: Faults
and folds. Journal of Geophysical Research: Solid Earth, 89(B13),
11517–11526. <https://doi.org/10.1029/JB089iB13p11517>

Mostafa, M. E. (2005). Iterative direct inversion: An exact
complementary solution for inverting fault-slip data to obtain
palaeostresses. Computers & Geosciences, 31(8), 1059–1070.
<https://doi.org/10.1016/j.cageo.2005.02.012>

Yamaji, A., & Sato, K. (2006). Distances for the solutions of stress
tensor inversion in relation to misfit angles that accompany the
solutions. Geophysical Journal International, 167(2), 933–942.
<https://doi.org/10.1111/j.1365-246X.2006.03188.x>
