---
title: "Fault Offsets"
author: "Tobias Stephan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
description: >
  Introduction into extracting offset components along faults
vignette: >
  %\VignetteIndexEntry{Fault Offsets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

This tutorial demonstrates how you
can derive displacement components from fault slip.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(20250411)
```


```{r load_functions, include=FALSE}
library("structr")
```

The offset along a fault can be factorized into several components.

![Fig. 1: Graphic illustration of displacement components along a fault.
$\mathbb{G}$ as the georeference frame with D = down, E = East, N =
North.](fault_displacements.png){#id .class width="75%" height="75%"}

## Get different components with trigonometry

### Fault orientation (dip angle, dip direction), shortening direction, and horizontal throw {#sec.offset.1}

Knowing the horizontal throw (e.g. from plate motion parameters), the
remaining components of the displacements along a given fault are as
follows.

\begin{equation}\delta = |\sigma_{\textrm{Hmax}} - (\textrm{dip direction}+90^{\circ})|\end{equation}

*Slip components in the horizontal plane:*

\begin{equation}\begin{split}
  f_\textrm{strike slip} & = |\cos{\delta} * f_\textrm{horizontal throw}|\\
  f_\textrm{heave} & = \sqrt{f_\textrm{horizontal throw}^2 - f_\textrm{strike slip}^2}
  \end{split}\end{equation}

*Slip components in the vertical plane perpendicular to the strike of
the fault:*

\begin{equation}\begin{split}
  f_\textrm{dip slip} &=  \frac{f_\textrm{heave}}{cos{(\textrm{dip})}}\\
  f_\textrm{vertical throw} &= \sqrt{f_\textrm{dip slip}^2 - f_\textrm{heave}^2}
  \end{split}\end{equation}

*Slip components in the fault plane plane:*

\begin{equation}\begin{split}
  f_\textrm{net slip} &= \sqrt{f_\textrm{strike slip}^2 + f_\textrm{dip slip}^2}\\
  \textrm{rake} &= \arctan{\left(\frac{f_\textrm{dip slip}}{f_\textrm{strike slip}}\right)}
  \end{split}\end{equation}

Thus, the rake angle describes the ratio between the dip slip and the
strike slip component.

Knowing the vertical throw (e.g. from thermochronology or petrology),
the fault dip (?assumption), and the direction and amount of horizontal
offset, the **strike** of the fault is as follows:

\begin{equation}\begin{split}
f_\textrm{heave} &= f_\textrm{vertical throw} * \tan{(\textrm{dip})}\\
\delta &= \arcsin{\left(\frac{f_\textrm{heave}}{f_\textrm{horizontal throw}}\right)}\\
\textrm{strike} &= |\sigma_{\textrm{Hmax}} - \delta|
\end{split}\end{equation}

Knowing the vertical throw (e.g. from thermochronology or petrology),
the fault strike (geomorphology), and the direction and amount of
horizontal offset, the **dip** of the fault is as follows:

\begin{equation}\begin{split}
\delta &= |\sigma_{\textrm{Hmax}} - \textrm{strike}|\\
f_\textrm{heave} &= f_\textrm{horizontal throw} * \sin{\delta} \\
\textrm{dip} &= \arctan{\left(\frac{f_\textrm{heave}}{f_\textrm{vertical throw}}\right)} 
\end{split}\end{equation}

Knowing the vertical throw (e.g. from thermochronology or petrology) and
the fault's dip and rake, the *horizontal offset*, the horizontal throw,
and the net-slip are as follows:

## Fault displacement tensors {#sec.tensors}

Each fault component is a vector describing its direction and length.
For instance, the vector of the strike slip is:

\begin{equation}
  \vec{f_\text{strike slip}} = \begin{pmatrix} \lVert f_\textrm{strike-slip}\rVert \\ 0 \\ 0 \end{pmatrix}
\end{equation}

## Net slip vector {#sec.tensors.netslip}

Net slip vector

\begin{equation}
  \vec{f_\text{net}} = \begin{pmatrix} \lVert f_\textrm{strike-slip}\rVert \\ \lVert f_\textrm{heave}\rVert \\ \lVert f_\textrm{vertical throw}\rVert \end{pmatrix}
\end{equation}

```{r netslip_vector, eval=FALSE}
fault_displacements(strikeslip = 2, verticalthrow = -5, heave = 3)
```

## Principal displacement tensor {#sec.tensors.principal}

The Eigen values of $\mathsf{F}_{ij}$; represented as
$\{ \vec{f_1}, \vec{f_2}, \vec{f_3} \}$ are referred to as the heave,
strike slip, and vertical throw component, respectively. These
orthonormal vectors define a orthogonal matrix, i.e. the **principal
displacement tensor** $\mathsf{F}_\mathbb{F}$:

\begin{equation}\mathsf{F}_{\mathbb{F}} = {\begin{bmatrix} \lVert f_\textrm{heave}\rVert & 0 & 0\\ 0 & \lVert f_\textrm{strike-slip}\rVert & 0\\ 0 & 0 & \lVert f_\textrm{vertical throw}\rVert\end{bmatrix}}\end{equation}

The tensor $\mathsf{F}$ can also be defined by the magnitudes of the
fault displacements:

```{r displacement_tensor}
Fu <- displacement_tensor(s = 2, v = -5, h = 3)
print(Fu)
```

## Orientation tensor {#sec.tensors.orientation}

Fault orientation tensor is defined by the fault plane's location,
orientation (dip direction and dip angle), and the fault's slip
(direction and magnitude):

\begin{equation}\mathsf{F}_{ij} = {\begin{bmatrix}f_{11} & f_{12} & f_{13}\\ f_{21} & f_{22} & f_{23}\\ f_{31} & f_{32} & f_{33}\end{bmatrix}}\end{equation}

```{r ortensor}
Fg <- displacement_tensor(s = 2, v = -5, h = 3, dip_direction = 45)
print(Fg)
```

## From Principal displacement tensor to Orientation tensor

Translation point of origin in $\mathsf{F_\mathbb{F}}$ into point of
measurement and rotate into fault orientation $\mathsf{F}_\mathbb{FG}$

```{r ftensor_decomp}
displacement_tensor_decomposition(Fg, dip_direction = 45)
```

