---
title: "10. Regression"
author: "Tobias Stephan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{10. Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

This tutorial demonstrates geodesic regression, a method to determine statistically the best-fit great- or small-circle for a set of vectors.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(20250411)
```

```{r setup, warning=FALSE, message=FALSE}
library(structr)
```

Import some example data of some folded fabric elements
```{r data, warning=FALSE, message=FALSE}
data("gray_example")
gray_example
```     
Show data in equal-area projection:

```{r plot_data, warning=FALSE, message=FALSE}
cleavage <- gray_example[1:8, ]
bedding <- gray_example[9:16, ]

stereoplot()
points(cleavage, col = "#1D1147")
points(bedding, col = "#E65164", pch = 4)

legend("topright", col = c("#1D1147", "#E65164"), pch = c(1, 4), legend = c("Poles to cleavage", "Poles to bedding"), bg = 'white')
``` 

The diagram suggests that both the cleavage and the bedding poles are oriented along great-circles or small-circles. To find out the orientation of these axes and decide whether the orientation better fit a great- or small-circle solution, we'll perform a (geodesic) regression.

## Best-fit great circles
First, we determine the best-fit great-circle for the two fabric elements. The great-circle regression is done via `regression_greatcircle()`, which a returns a list containing the axis of the great-circle:

```{r calc_greatcircle, warning=FALSE, message=FALSE}
cleavage_gc <- regression_greatcircle(cleavage)
bedding_gc <- regression_greatcircle(bedding)

# The axis of the great circles:
cleavage_gc$vec
bedding_gc$vec
```

The algorithm minimizes the sum-of-squares of the residuals, which are the angular distances of the best-fit axes to the data vectors. 
We can check if the optimization successfully converged, and asses the quality of the fit using the R<sup>2</sup>:
```{r check_greatcircle, warning=FALSE, message=FALSE}
# Check 
cleavage_gc$convergence == 0 & cleavage_gc$min_eigenvalue > 0
bedding_gc$convergence == 0 & bedding_gc$min_eigenvalue > 0

# R^2 value
cleavage_gc$r_squared
bedding_gc$r_squared
``` 

Plot the results
```{r plot_greatcircle, warning=FALSE, message=FALSE}
cleavage_gc <- regression_greatcircle(cleavage)
bedding_gc <- regression_greatcircle(bedding)

stereoplot()
lines(cleavage_gc$vec, col = "#000004FF")
lines(bedding_gc$vec, col = "#B63679")
points(cleavage, col = "#1D1147")
points(bedding, col = "#E65164", pch = 4)

legend("bottomright", col = c("#000004FF", "#B63679"), lty =c(1, 1), legend = c("Cleavage best greatcircle", "Bedding best greatcircle"), bg = 'white')
``` 

### Summary

The regression was successful. However, the R<sup>2</sup> values indicate a good fit for the bedding, but a rather poor fit for the cleavage planes.

## Best-fit small circles
```{r calc_smallcircle, warning=FALSE, message=FALSE}
cleavage_sc <- regression_smallcircle(cleavage)
bedding_sc <- regression_smallcircle(bedding)

# Half-apical angle of small-circle cone (in degrees)
cleavage_sc$cone
bedding_sc$cone
``` 

Check the optimization convergence and the fit:
```{r check_smallcircle, warning=FALSE, message=FALSE}
# Check 
cleavage_sc$convergence == 0 & cleavage_sc$min_eigenvalue > 0
bedding_sc$convergence == 0 & bedding_sc$min_eigenvalue > 0

# R^2 value
cleavage_sc$r_squared
bedding_sc$r_squared
``` 

Plot the results:
```{r plot_smallcircle, warning=FALSE, message=FALSE}
stereoplot()
lines(cleavage_sc$vec, cleavage_sc$cone, col = "#000004FF")
lines(bedding_sc$vec, bedding_sc$cone, col = "#B63679")
points(cleavage, col = "#1D1147")
points(bedding, col = "#E65164", pch = 4)

legend("bottomright", col = c("#000004FF", "#B63679"), lty =c(1, 1), legend = c("Cleavage best smallcircle", "Bedding best smallcircle"), bg = 'white')
```

### Summary

The regression was successful. The R<sup>2</sup> values indicate a better fit for the cleavage planes, but a slightly poorer fit for the bedding planes. Hence, the poles to the bedding planes seem to be aligned along a great-circle, while the cleavage planes better fit a small-circle arrangement.

# References
Bachmann, F., Jupp, P. E., & Schaeben, H. (2014). Estimating the number and locations of Euler poles. International Journal on Geomathematics, 5(2), 289–301. https://doi.org/10.1007/s13137-014-0064-2

Davis, J. R., & Titus, S. J. (2017). Modern methods of analysis for three-dimensional orientational data. Journal of Structural Geology, 96, 65–89. https://doi.org/10.1016/j.jsg.2017.01.002

Gray, N. H., Geiser, P. A., & Geiser, J. R. (1980). On the least-squares fit of small and great circles to spherically projected orientation data. Journal of the International Association for Mathematical Geology, 12(3), 173–184. https://doi.org/10.1007/BF01091203

Ramsay, J. G. (1967). Folding and Fracturing of Rocks. McGraw-Hill.