---
title: "Statistics on the Sphere"
author: "Tobias Stephan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
description: >
  Some basic staistical analyses tools for spherical data.
vignette: >
  %\VignetteIndexEntry{Statistics on the Sphere}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

This tutorial describes how to calculate basic statistics for spherical
orientation data using the `{structr}` package. It covers the
calculation of mean orientations, dispersion measures, hypothesis
testing, orientation tensors, and clustering of orientation data.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(20250411)
```

Load the package:

```{r setup, warning=FALSE, message=FALSE}
library(structr)
```

To start, we import some example planes and lines, and convert them to
spherical objects:

```{r example}
data(example_planes_df)
data(example_lines_df)

planes <- Plane(example_planes_df$dipdir, example_planes_df$dip)
lines <- Line(example_lines_df$trend, example_lines_df$plunge)
```

## Mean and dispersion

Arithmetic mean, geodesic mean, and projected mean

### Arithmetic mean and variance

The arithmetic mean orientation of spherical data is calculated by
summing up all orientation vectors and normalizing the resulting
vector[^1]:

[^1]: Mardia, K. V., & Jupp, P. E. (2000). Directional Statistics. (K.
    V. Mardia & P. E. Jupp, Eds.). Hoboken, NJ, USA: Wiley.
    <https://doi.org/10.1002/9780470316979>

This involves the resultant vector of a set of $n$ vectors $x_1, \ldots, x_n$, given by $\bar{\mathbf{x}} = \sum_{i=1}^{n} \mathbf{x}_i$,
and the resultant length defined as $\bar{\mathbf{R}} = \sqrt{x_x^2 + x_y^2 + x_z^2}$.

The arithmetic mean of the vector components is the mean resultant vector, i.e the resultant vector normalized by its length
$$\mu = \frac{\bar{\mathbf{x}}}{\bar{\mathbf{R}}}$$

In geology, the statistical estimators are useful for **vectorial data**
(`"Line"` objects) such as lineations (e.g. mineral lineations,
stretching lineations) or plunge directions of fold axes, but less for
axial data (`"Ray"` objects) such as paleomagnetic directions or
orientations of fold axes, where the direction is not relevant[^2].

[^2]: Bachmann, F., Hielscher, R., Jupp, P. E., Pantleon, W., Schaeben,
    H., & Wegert, E. (2010). Inferential statistics of electron
    backscatter diffraction data from within individual crystalline
    grains. Journal of Applied Crystallography, 43(6), 1338–1355.
    <https://doi.org/10.1107/S002188981003027X>

These estimators are good descriptors of the concentration of the data
around the mean direction, assuming an unimodal, isotropic distributions
such as the *von Mises-Fisher distribution*.

Define some weights for our lines based on the quality reported for the
measurements:

```{r stat0}
example_lines_df$quality <- ifelse(is.na(example_lines_df$quality), 6, example_lines_df$quality) # replacing NA values with 6
line_weightings <- 6 / example_lines_df$quality
```

The (weighted) arithmetic mean orientation of spherical data is:

```{r stat_mean}
lines_mean <- sph_mean(lines, w = line_weightings)
```

... and the (weighted) arithmetic variance

```{r stat_var}
lines_variance <- sph_var(lines, w = line_weightings)
```

the (weighted) standard deviation (i.e. the 63% cone around the mean)
and the 95% confidence cone around the mean:

```{r stat_delta}
lines_delta <- delta(lines, w = line_weightings)
lines_confangle <- sph_confidence_angle(lines, w = line_weightings)
```

Taken together, this prints as

```{r stat}
c(
  "Variance" = lines_variance,
  "63% cone" = lines_delta,
  "Confidence angle" = lines_confangle
)
```

Summary stats can also be retrieved through

```{r summary}
summary(lines)
```

Plotting a summary of the stats on a equal-area projection:

```{r stat_plot, message=FALSE,warning=FALSE}
#| fig.alt: >
#|   Diagram showing statistical results of some example data plotted in an
#|   equal-area projection
stereoplot()
points(lines, col = "lightgrey", pch = 1, cex = .5)
points(lines_mean, col = "#B63679", pch = 19, cex = 1)
lines(lines_mean, ang = lines_confangle, col = "#E65164FF")
lines(lines_mean, ang = lines_delta, col = "#FB8861FF")
legend("topright", legend = c("Mean line", "95% confidence cone", "63% data cone"), col = c("#B63679", "#E65164FF", "#FB8861FF"), pch = c(19, NA, NA), lty = c(NA, 1, 1), cex = .75)
```

### Geodesic mean and variance

Another measure of mean and dispersion is the *Fréchet (geodesic
L*<sup>2</sup>) mean and *variance* which is based on the angles between
all data vectors[^3].\
This can be visualized using the `variance_plot()`; here showing the
(geodesic) angle distances between all lines and the first line in the
same data set:

[^3]: Davis, J. R., & Titus, S. J. (2017). Modern methods of analysis
    for three-dimensional orientational data. Journal of Structural
    Geology, 96, 65–89. <https://doi.org/10.1016/j.jsg.2017.01.002>

```{r variance}
#| fig.alt: >
#|   Diagram showing the distances of a set of vectors to a specified vector
variance_plot(lines, y = lines[1, ])
```

> The Fréchet mean is the vector that minimizes the sum of the squared
> angle distances to all data vectors. The Fréchet variance about this
> mean is the sum of the squared angle distances between all data
> vectors and the Fréchet mean.

To find the Fréchet mean vector, the algorithm iteratively searches for
the vector by using a numerical optimization method to minimize the
distances to all other vectors

```{r geodesic}
# Mean
geodesic_mean(lines)

# Variance
geodesic_var(lines)
```

These estimators are good descriptors of the concentration of the data
around the mean direction, when data follows a unimodal, anisotropic
distributions such as the *Kent* or *Bingham distributions*.

### Projected mean

The Eigenvector with the largest Eigenvalue represents a vector parallel
to the highest concentration of a population. This vector can also be
described as the **projected mean**. A shortcut function for this is:

```{r projected}
projected_mean(example_lines)
```

These estimators are good descriptors of the concentration of the data
around the mean direction, when data follows an unimodal, anisotropic
distributions such as the *Kent* or *Bingham distributions*.

```{r mean_plot}
#| fig.alt: >
#|  Equal area projections showing the different estimators of the mean

par(mfrow = c(1, 2))
stereoplot()
points(example_lines, col = "grey", pch = 16, cex = .7)
points(sph_mean(example_lines), col = "#1D1147FF", pch = 16)
points(geodesic_mean(example_lines), col = "#B63679FF", pch = 16)
points(projected_mean(example_lines), col = "#FEC287FF", pch = 16)
title(main = "Lines")

legend("bottomleft",
  legend = c("Arithmetic mean", "Geodesic mean", "Projected mean"),
  col = c("#1D1147FF", "#B63679FF", "#FEC287FF"),
  pch = 16,
  bg = "white"
)

stereoplot()
points(example_planes, col = "grey", pch = 16, cex = .7)
points(sph_mean(example_planes), col = "#1D1147FF", pch = 16)
points(geodesic_mean(example_planes), col = "#B63679FF", pch = 16)
points(projected_mean(example_planes), col = "#FEC287FF", pch = 16)
title(main = "Planes")
```

### Hypothesis testing

To test if a line represents the population mean of a given set of
lineations, we need to calculate the confidence region of our
population.

Let's test the hypothesis that a horizontal lineation trending towards
70° is the mean for our lineations.

```{r test0}
line_NULL <- Line(70, 0)
```

The 95% confidence interval (from 10,000 bootstrap samples):

```{r test1}
ce <- confidence_ellipse(lines, n = 10000, alpha = 0.05)
```

To visualize the confidence region of our lines:

```{r test2}
#| fig.alt: Diagram showing the confidence region of a mean
plot(lines, col = "grey")
stereo_confidence(ce, col = "#B63679FF")
points(line_NULL, col = "#000004", pch = 16)
```

The p-value for our hypothesis line:

```{r test3}
ce$pvalue.FUN(line_NULL)
```

With (95% confidence) we rejected the Null Hypothesis that the given
line represents the population mean as the p-value is smaller than 5%.

## Orientation tensor

### Eigenvectors

The orientation tensor[^4] is a matrix comprising the mean direction
cosines of the orientation vectors. In case of a Bingham distribution,
the **Eigenvectors** of this tensor describe the orientation of the most
dense, intermediate and least dense orientation, and thus, are used to
determine the orientation of girdle-distributed vectors (e.g. folded
planes).

[^4]: Scheidegger, A. E. (1964). The tectonic stress and tectonic motion
    direction in Europe and Western Asia as calculated from earthquake
    fault plane solutions. Bulletin of the Seismological Society of
    America, 54(5A), 1519–1528. <https://doi.org/10.1785/BSSA05405A1519>

#### Shape parameters

There are more shape parameters using different algorithms based on the
orientation tensor:

```{r shapes}
shape_params(planes)
```

## Cluster vectors

To find k clusters of orientational data, the `sph_cluster()` function
can be used:

```{r cluster}
#| fig.alt: >
#|   Stereoplot showing the the cluster result of an example dataset

# generate some random vectors:
x1 <- rvmf(100, mu = Line(90, 0), k = 20)
x2 <- rvmf(100, mu = Line(0, 0), k = 20)
x3 <- rvmf(100, mu = Line(0, 90), k = 20)
x123 <- rbind(x1, x2, x3)

# cluster the vectors:
cl <- sph_cluster(x123, k = 3)

# visualize the result:
plot(x123, col = assign_col_d(cl$cluster))

legend_d(assign_col_d(unique(cl$cluster)), title = "Cluster")
```

## References

Bachmann, F., Hielscher, R., Jupp, P. E., Pantleon, W., Schaeben, H., &
Wegert, E. (2010). Inferential statistics of electron backscatter
diffraction data from within individual crystalline grains. Journal of
Applied Crystallography, 43(6), 1338–1355.
<https://doi.org/10.1107/S002188981003027X>

Davis, J. R., & Titus, S. J. (2017). Modern methods of analysis for
three-dimensional orientational data. Journal of Structural Geology, 96,
65–89. <https://doi.org/10.1016/j.jsg.2017.01.002>

Mardia, K. V., & Jupp, P. E. (2000). Directional Statistics. (K. V.
Mardia & P. E. Jupp, Eds.). Hoboken, NJ, USA: Wiley.
<https://doi.org/10.1002/9780470316979>

Scheidegger, A. E. (1964). The tectonic stress and tectonic motion
direction in Europe and Western Asia as calculated from earthquake fault
plane solutions. Bulletin of the Seismological Society of America,
54(5A), 1519–1528. <https://doi.org/10.1785/BSSA05405A1519>
